### 2023-12-15 submission/manuscript.Rmd
Final polish before submitting

Considering it is common to find as much as 100-fold variation in the number of 16S rRNA gene sequences found across samples in a study, researchers need to control for the effect of uneven sequencing effort. How to do this has become a contentious question. Some have argued that rarefying or rarefaction is "inadmissible" because it omits valid data. A number of alternative approaches have been developed to normalize and rescale the data that purport to be invariant to the number of observations. I generated community distributions based on 12 published datasets where I was able to assess the ability of multiple methods to control for uneven sequencing effort. Rarefaction was the only method that could control for variation in uneven sequencing effort when measuring commonly used alpha and beta diversity metrics. Next, I compared the false detection rate and power to detect true differences between simulated communities with a known effect size using various alpha and beta diversity metrics. Although all methods of controlling for uneven sequencing effort had an acceptable false detection rate when samples were randomly assigned to two treatment groups, rarefaction was consistently able to control for differences in sequencing effort when sequencing depth was confounded with treatment group. Finally, the statistical power to detect differences in alpha and beta diversity metrics was consistently the highest when using rarefaction. These simulations underscore the importance of using rarefaction to normalize the number of sequences across samples in amplicon sequencing analyses.

>>>>

Considering it is common to find as much as 100-fold variation in the number of 16S rRNA gene sequences across samples in a study, researchers need to control for the effect of uneven sequencing effort. How to do this has become a contentious question. Some have argued that rarefying or rarefaction is "inadmissible" because it omits valid data. A number of alternative approaches have been developed to normalize and rescale the data that purport to be invariant to the number of observations. I generated community distributions based on 12 published datasets where I was able to assess the ability of multiple methods to control for uneven sequencing effort. Rarefaction was the only method that could control for variation in uneven sequencing effort when measuring commonly used alpha and beta diversity metrics. Next, I compared the false detection rate and power to detect true differences between simulated communities with a known effect size using various alpha and beta diversity metrics. Although all methods of controlling for uneven sequencing effort had an acceptable false detection rate when samples were randomly assigned to two treatment groups, rarefaction was consistently able to control for differences in sequencing effort when sequencing depth was confounded with treatment group. Finally, the statistical power to detect differences in alpha and beta diversity metrics was consistently the highest when using rarefaction. These simulations underscore the importance of using rarefaction to normalize the number of sequences across samples in amplicon sequencing analyses.

### 2023-12-15 submission/manuscript.tex
Final polish before submitting

Considering it is common to find as much as 100-fold variation in the
number of 16S rRNA gene sequences found across samples in a study,
researchers need to control for the effect of uneven sequencing effort.
How to do this has become a contentious question. Some have argued that
rarefying or rarefaction is ``inadmissible'' because it omits valid
data. A number of alternative approaches have been developed to
normalize and rescale the data that purport to be invariant to the
number of observations. I generated community distributions based on 12
published datasets where I was able to assess the ability of multiple
methods to control for uneven sequencing effort. Rarefaction was the
only method that could control for variation in uneven sequencing effort
when measuring commonly used alpha and beta diversity metrics. Next, I
compared the false detection rate and power to detect true differences
between simulated communities with a known effect size using various
alpha and beta diversity metrics. Although all methods of controlling
for uneven sequencing effort had an acceptable false detection rate when
samples were randomly assigned to two treatment groups, rarefaction was
consistently able to control for differences in sequencing effort when
sequencing depth was confounded with treatment group. Finally, the
statistical power to detect differences in alpha and beta diversity
metrics was consistently the highest when using rarefaction. These
simulations underscore the importance of using rarefaction to normalize
the number of sequences across samples in amplicon sequencing analyses.

>>>>

Considering it is common to find as much as 100-fold variation in the
number of 16S rRNA gene sequences across samples in a study, researchers
need to control for the effect of uneven sequencing effort. How to do
this has become a contentious question. Some have argued that rarefying
or rarefaction is ``inadmissible'' because it omits valid data. A number
of alternative approaches have been developed to normalize and rescale
the data that purport to be invariant to the number of observations. I
generated community distributions based on 12 published datasets where I
was able to assess the ability of multiple methods to control for uneven
sequencing effort. Rarefaction was the only method that could control
for variation in uneven sequencing effort when measuring commonly used
alpha and beta diversity metrics. Next, I compared the false detection
rate and power to detect true differences between simulated communities
with a known effect size using various alpha and beta diversity metrics.
Although all methods of controlling for uneven sequencing effort had an
acceptable false detection rate when samples were randomly assigned to
two treatment groups, rarefaction was consistently able to control for
differences in sequencing effort when sequencing depth was confounded
with treatment group. Finally, the statistical power to detect
differences in alpha and beta diversity metrics was consistently the
highest when using rarefaction. These simulations underscore the
importance of using rarefaction to normalize the number of sequences
across samples in amplicon sequencing analyses.
